---
title: "01 - Introduction to Computing using R"
subtitle: "Junvie Pailden"   
author: "SIUE, F2017, Stat 589"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  beamer_presentation:
    theme: "Boadilla"
    colortheme: "seagull"
    fonttheme: "serif"
  #ioslides_presentation: default
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  comment = "#",
  fig.path = "figures/",
  fig.height = 3.5,
  fig.width = 3.5,
  fig.align = "center",
  cache = FALSE,
  warnings = FALSE,
  message = FALSE)
knitr::opts_knit$set(
  root.dir="data/"
  )
```

## Why R

- `R` offers a powerful and appealing interactive environment for exploring data, running simulations, etc.  

\itemsep0em

- `R` is platform independent meaning it is available on Windows, Mac, and Linux.  

\itemsep0em

- `R`` has excellent resources both online (just google any issue/question) and using help(...), e.g. `help(lm)`.  

\itemsep0em

- `R` is not a black box software, i.e., you can trace how a function or package works by following the {\tt R} script, e.g. `lm()`


## R as a calculator

```{r}
# create an integer sequence 
3:7 
# create an sequence from 0 to 3 with 0.5 increment 
seq(0,3,by=0.5) 
# create a repeated sequence 
rep(pi,4)
# Basic Operations
(2-3)*(3/4)
```


## Concatenate operator

```{r}
c(6,20,-3) # numbers 
c(log(100), log10(100))
c("words","are","wind") # strings 
```

## Variable assignment, common operator

```{r}
a <- 3 
# is the same as 
a -> x 
# and x = 3
b <- 1:3; c <- 2:4
a + b # scalar multiplication
b * c # entrywise multiplication
```


## Matrices

```{r}
# store the vector (1, 2, 3, 4, 5, 6, 7, 8) 
# into a 2x4 matrix named A
# nrow = # of rows, ncol = # of columns
A <- matrix(c(1,2,3,4,5,6,7,8), nrow = 2, ncol = 4) 
colnames(A) <- c('C1', 'C2', 'C3', 'C4')
A
B <- matrix(1:8, nrow = 2, ncol = 4, byrow = TRUE) # 1:8 is another way of generating  the vector
B
```

## Extracting matrix entries

```{r}
A[1, 3]   # row 1, col 3 entry
A[1, 1:3] # row 1, col 1 to 3
A[1, ]    # row 1
A[ , 3]   # col 3
```

## Data Frames

Data frames are similar in concept to Excel spreadsheet where each column represents a variable or measurement and each row represents and observational unit. 

We can check the structure of the data frame using the function `str()`. The output includes data information such as

* dimension (number of observations and variables), 
* variable names
* data type 
    - `logi`: logical, `TRUE` or `FALSE`
    - `int`: integer
    - `num`: numerical 
    - `chr`: character
    - `Factor`: Factors are how R keeps track of categorical variables.


## 

Use `data.frame()`:

```{r}
stark.kids <- data.frame(
  Name = c("Jon", "Sansa", "Arya", "Bran"),
  Age = c(24, 20, 18, 17)
)
str(stark.kids)
stark.kids
```

## Variable names

We can also use the names of variables in a data frame to access the variable using the notation `data$name`.

```{r }
stark.kids$Name
mean(stark.kids$Age) # average age
head(stark.kids, 2) # display first 2 rows
```


## Reading data in R: Example 1.6 Lizard Size Data

Data analysis using R often involves importing or reading data at some point.

```{r}
# File "T1-3.dat"" should be in your working directory
# Specify the working directory
# Session > Set Working Directory > Choose Directory
lizard <- read.table("T1-3.dat")
# assign names to the column variables
colnames(lizard) <- c("mass", "svl", "hls") 
head(lizard, 3) # display first 3 rows
```

## Reading CSV data in R

`Goind Wireless` data reported the estimated percentage of house- holds with only wireless phone service (no land line) for the 50 U.S. states and the District of Columbia.

The URL of the data set `Going Wireless` that we need to read the data is (<https://goo.gl/72BKSf>).

```{r 16-wk02}
wireless.data <- read.csv("https://goo.gl/72BKSf", 
                          header = TRUE)
str(wireless.data) # check structure
```

## Simple data summaries

```{r}
height <- 58:72
weight <- c(115,117,120,123,126,129,132,135,139,
				142,146,150,154,159,164)
mean(height) # mean of height 
var(height)  # variance of height 
length(height)
cor(height, weight) # correlation
```

## Writing Functions in R

```{}
function.name <- function(arglist){
    expr
}
```

```{r}
my_fun <- function( x, y ){
	  x + y 
}
my_fun(1,2)
my_fun2 <- function(x, y) x*y
my_fun2(10, 23)
```
If the body of the function is only one line, then braces aren't necessary.


## `apply()` function

Applies a function to sections of an array (or matrix) and returns the results in an array (or matrix). 

```{r}
# create a matrix of twenty rows, two columns
dat <- cbind(rbinom(n = 20, size = 5, prob = 0.3), 
             rnorm(n = 20)) 
str(dat)
apply(dat, MARGIN = 2, var) # apply variance to columns
```


## Conditionals

```{}
if (arglist is satisfied) {   
	do this 
} else {   
	do this instead
}
```
```{r}
my_cond <- function( x ){ 
  # function that tells whether a value is > 20.
  if (x > 20) {
	  print("x is greater than 20") 
  } else {   
	  print("x is less than 20") 
  } 
} 
my_cond(10)
```

## Loops in `R`

For loops iterate through each item in a vector or a list:

```{}
for (x in vector) {
  do this
}
```
The colon creates a vector, passing each integer from 0 to 4 to the loop.
```{r}
for (x in 0:4) print(x)
```
Two more loops in `R` are `repeat()` and `while()`. Search this!


## Packages in R! Treasure troves of goodies

- An `R` package is a set of related functions and help files, bundled
together.  

- Normally, all functions within a single package are related: for example, the  `stats` package contains functions for statistical analysis.  

- There are few public repositories of packages: the largest is CRAN hosted by the `R` foundation with more than 4000 freely packages.  

- To use a package, you first need to install it into `R`.  
- Among other ways, you can also install `R` packages directly through `R` console using `install.packages()`.   

- To load up an `R` package, use the `library()`. 


## Learn More

1. [An Introduction to R by WN Venables](https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf)

\itemsep0em

2. [Using R Markdown for Class Reports](http://www.stat.cmu.edu/~cshalizi/rmarkdown/)

\itemsep0em

3. [Quick R website](http://www.statmethods.net/)

\itemsep0em

4. [Online learning by RStudio](https://www.rstudio.com/online-learning/)

\itemsep0em

5. [R for Data Science by H Wickham](http://r4ds.had.co.nz/)

\itemsep0em

6. Just google it!