---
title: "10 - Multivariate Analysis of Variance"
subtitle: "Junvie Pailden"   
author: "SIUE, F2017, Stat 589"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
    beamer_presentation:
      theme: "Singapore"
      colortheme: "lily"
      fonttheme: "professionalfonts"
  #ioslides_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  options(digits = 2), # round to 2 digits
  echo = TRUE,
  comment = "#",
  fig.path = "figures/",
  fig.height = 3.5,
  fig.width = 3.5,
  fig.align = "center",
  cache = FALSE,
  warnings = FALSE,
  message = FALSE)
knitr::opts_knit$set(
  root.dir="data/"
  )
```


## Comparing Several Mult Pop'n Means (Multivariate ANOVA)

Random samples, collected from each of $g$ populations,

\begin{table}
\centering
\begin{tabular}{|c|c|}
\hline 
Population 1: & $\mathbf{X}_{11},\mathbf{X}_{12},\ldots,\mathbf{X}_{1n_{1}}$\tabularnewline
\hline 
Population 2: & $\mathbf{X}_{21},\mathbf{X}_{22},\ldots,\mathbf{X}_{2n_{2}}$\tabularnewline
\hline 
$\vdots$ & $\vdots$\tabularnewline
\hline 
Population $g$: & $\mathbf{X}_{g1},\mathbf{X}_{g2},\ldots,\mathbf{X}_{gn_{g}}$\tabularnewline
\hline 
\end{tabular}
\end{table}

* MANOVA is used first to investigate whether the population mean vectors are the same and, if not, which mean components differ significantly.


## Review of Univariate ANOVA

* $X_{\ell1},X_{\ell2},\ldots,X_{\ell n_{l}}$ is a random sample from an $N(\mu_{\ell},\sigma^{2})$ population, $\ell=1,2,\ldots,g$  

* random samples are independent  

* $H_{0}:\mu_{1}=\mu_{2}=\cdots=\mu_{q}$  

* $\mu_{\ell}=\mu+(\mu_{\ell}-\mu)=\mu+\tau_{\ell}$, where $\tau_{\ell}=\mu_{\ell}-\mu$.

* The null hypothesis becomes $H_{0}:\tau_{1}=\tau_{2}=\cdots=\tau_{g}=0$  

* The response $X_{\ell j}\sim N(\mu+\tau_{j},\sigma^{2})$, can be written as
\[
X_{\ell j}=\mu+\tau_{\ell}+e_{\ell j}
\]
where the $e_{\ell j}$ are independent $N(0,\sigma^{2})$ random variables.

* To define uniquely the model parameters and their least squares estimates, we impose the constraint
\[
\sum_{\ell=1}^{g}n_{\ell}\tau_{\ell}=0.
\]


## 

The analysis of variance is based upon an analogous decomposition of the observations
\begin{align*}
x_{\ell j} & =\bar{x}+(\bar{x}_{\ell}-\bar{x})+(x_{\ell j}-\bar{x}_{\ell})\\
 & =\bar{x}+\hat{\tau}_{\ell}+\hat{e}_{\ell j}\\
\text{(obs)} & =\text{(overall sample mean)}+ \\
& \quad \text{(estimated treatment effect)}+\text{(residual)}
\end{align*}

Note that, for all $\ell=1,2,\ldots,g$,
\[
\sum_{\ell=1}^{n_{\ell}}(x_{\ell j}-\bar{x})^{2}=n_{\ell}(\bar{x}_{\ell}-\bar{x})^{2}+\sum_{j=1}^{n_{\ell}}(x_{\ell j}-\bar{x}_{\ell})^{2},
\]
since $\sum_{j=1}^{n_{\ell}}(x_{\ell j}-\bar{x}_{\ell})=0$.



## 

Summing both sides over $\ell$ we get
\begin{align*}
\sum_{\ell=1}^{g}\sum_{j=1}^{n_{\ell}}(x_{\ell j}-\bar{x})^{2} & =\sum_{\ell=1}^{g}n_{\ell}(\bar{x}_{\ell}-\bar{x})^{2}+\sum_{\ell=1}^{g}\sum_{j=1}^{n_{\ell}}(x_{\ell j}-\bar{x}_{\ell})^{2}\\
\left(\begin{array}{c}
SS_{cor}\\
\text{total}\\
\text{(corrected) SS}
\end{array}\right) & =\left(\begin{array}{c}
SS_{tr}\\
\text{between}\\
\mbox{\text{(samples) SS}}
\end{array}\right)+\left(\begin{array}{c}
SS_{res}\\
\text{within}\\
\text{(samples) SS }
\end{array}\right) \\
\text{OR} \qquad & \\
\sum_{\ell=1}^{g}\sum_{j=1}^{n_{\ell}}x_{\ell j}^{2}  &=(n_{1}+n_{2}+\cdots+n_{g})\bar{x}^{2}+\sum_{\ell=1}^{g}n_{\ell}(\bar{x}_{\ell}-\bar{x})^{2} \\
& \qquad + \sum_{\ell=1}^{g}\sum_{j=1}^{n_{\ell}}(x_{\ell j}-\bar{x}_{\ell})^{2}\\
(SS_{obs}) & =(SS_{mean})+(SS_{tr})+(SS_{res})
\end{align*}


## ANOVA Table

\begin{table}
\centering
\begin{tabular}{|c|c|c|}
\hline 
$\begin{array}{c}
\text{Source of}\\
\text{variation}
\end{array}$ & $\begin{array}{c}
\text{Sum of}\\
\text{Squares} (SS)
\end{array}$ & $\begin{array}{c}
\text{Degrees of}\\
\text{freedom} (d.f.)
\end{array}$\tabularnewline
\hline 
\hline 
Treatments & $SS_{tr}=\sum_{\ell=1}^{g}n_{\ell}(\bar{x}_{\ell}-\bar{x})^{2}$ & $g-1$\tabularnewline
\hline 
$\begin{array}{c}
\text{Residual}\\
\text{(error)}
\end{array}$ & $SS_{res}=\sum_{\ell=1}^{g}\sum_{j=1}^{n_{\ell}}(x_{\ell j}-\bar{x}_{\ell})^{2}$  & $\sum_{\ell=1}^{g}n_{\ell}-g$\tabularnewline
\hline 
$\begin{array}{c}
\text{Total}\\
\text{(corrected for}\\
\text{the mean)}
\end{array}$ & $SS_{cor}=\sum_{\ell=1}^{g}\sum_{j=1}^{n_{\ell}}(x_{\ell j}-\bar{x})^{2}$ & $\sum_{\ell=1}^{g}n_{\ell}-1$\tabularnewline
\hline 
\end{tabular}
\end{table}


## ANOVA Test for Comparing Univariate Means

* The usual $F$-test rejects $H_{0}:\tau_{1}=\tau_{2}=\cdots=\tau_{g}=0$
at level $\alpha$ if
\[
F=\frac{SS_{tr}/(g-1)}{SS_{res}/\left(\sum_{\ell=1}^{g}n_{\ell}-g\right)}>F_{g-1,\sum n_{\ell}-g}(\alpha)
\]
where $F_{g-1,\sum n_{\ell}-g}(\alpha)$ is the upper $(100\alpha)$th percentile of the $F$-distribution with $g-1$ and $\sum n_{\ell}-g$
degrees of freedom.  

* This is equivalent to rejecting $H_{0}$ for large values of $SS_{tr}/SS_{res}$ or for large values of $1+SS_{tr}/SS_{res}$.  

* The multivariate generalization rejects $H_{0}$ for small values of the reciprocal
\[
\frac{1}{1+SS_{tr}/SS_{res}}=\frac{SS_{res}}{SS_{res}+SS_{tr}}
\]


## Multivariate Analysis of Variance (MANOVA)

* MANOVA Model for Comparing $g$ Population Mean Vectors
\[
\mathbf{X}_{\ell j}=\mu+\tau_{\ell}+\mathbf{e}_{\ell j},\,\,\,j=1,2,\ldots,n_{\ell}\,\,\text{and}\,\,\ell=1,2,\ldots,g
\]
where the $\mathbf{e}_{\ell j}$ are independent $N_{p}(0,\Sigma)$ variables.   

* The parameter vector $\mu$ is an overall mean (level), and $\tau_{\ell}$ represents the $\ell$th treatment effect with
$\sum_{\ell=1}^{g}n_{\ell}\tau_{\ell}=0$.


## 

A vector of observations may be decomposed 

\begin{align*}
\mathbf{x}_{\ell j} & =\bar{\mathbf{x}}+(\bar{\mathbf{x}}_{\ell}-\bar{\mathbf{x}})+(\mathbf{x}_{\ell j}-\bar{\mathbf{x}}_{\ell})\\
\text{(observation)} & =\left(\begin{array}{c}
\text{overall sample}\\
\text{mean }\,\hat{\mu}
\end{array}\right)+\text{\ensuremath{\left(\begin{array}{c}
 \text{estimated }\\
 \text{treatment}\\
 \text{effect}\,\,\hat{\tau}_{\ell} 
\end{array}\right)}}+\text{\ensuremath{\left(\begin{array}{c}
 \text{residual}\\
 \hat{\mathbf{e}}_{\ell j} 
\end{array}\right)}}
\end{align*}


## 

Similarly, we have
\begin{align*}
\sum_{\ell=1}^{g}\sum_{j=1}^{n_{\ell}}(\mathbf{x}_{\ell j}-\bar{\mathbf{x}})(\mathbf{x}_{\ell j}-\bar{\mathbf{x}})' & =\sum_{\ell=1}^{g}n_{\ell}(\bar{\mathbf{x}}_{\ell}-\bar{\mathbf{x}})(\bar{\mathbf{x}}_{\ell}-\bar{\mathbf{x}})'+\sum_{\ell=1}^{g}\sum_{j=1}^{n_{\ell}}(\mathbf{x}_{\ell j}-\bar{\mathbf{x}}_{\ell})(\mathbf{x}_{\ell j}-\bar{\mathbf{x}}_{\ell})'\\
\left(\begin{array}{c}
SS_{cor}\\
\text{total}\\
\text{(corrected) SS}
\end{array}\right) & =\left(\begin{array}{c}
SS_{tr}\\
\text{between}\\
\mbox{\text{(samples) SS}}
\end{array}\right)+\left(\begin{array}{c}
SS_{res}\\
\text{within}\\
\text{(samples) SS }
\end{array}\right)\\
 & =\mathbf{B}+\mathbf{W}
\end{align*}


## 

The within sum of squares and cross products matrix can be expressed
as
\begin{align*}
\mathbf{W} & =\sum_{\ell=1}^{g}\sum_{j=1}^{n_{\ell}}(\mathbf{x}_{\ell j}-\bar{\mathbf{x}}_{\ell})(\mathbf{x}_{\ell j}-\bar{\mathbf{x}}_{\ell})'\\
 & =(n_{1}-1)\mathbf{S}_{1}+(n_{2}-1)\mathbf{S}_{2}+\cdots+(n_{g}-1)\mathbf{S}_{g}
\end{align*}
where $\mathbf{S}_{\ell}$ is the sample covariance matrix for the $\ell$th sample.


## MANOVA Table

The hypothesis of no treatment effects
\[
H_{0}:\tau_{1}=\tau_{2}=\cdots=\tau_{g}=0
\]
is tested by considering the relative sizes of the treatment and residual sums of squares and cross products.

\begin{table}
\centering

\begin{tabular}{|c|c|}
\hline 
$\begin{array}{c}
\text{Matrix of sum of squares and}\\
\text{cross products}(SSP)
\end{array}$ & $\begin{array}{c}
\text{Degrees of}\\
\text{freedom}(d.f.)
\end{array}$\tabularnewline
\hline 
\hline 
Treatments: \qquad $\mathbf{B}=\sum_{\ell=1}^{g}n_{\ell}(\bar{\mathbf{x}}_{\ell}-\bar{\mathbf{x}})(\bar{\mathbf{x}}_{\ell}-\bar{\mathbf{x}})'$ & $g-1$\tabularnewline
\hline 
Residual: \quad 
$\mathbf{W}=\sum_{\ell=1}^{g}\sum_{j=1}^{n_{\ell}}(\mathbf{x}_{\ell j}-\bar{\mathbf{x}}_{\ell})(\mathbf{x}_{\ell j}-\bar{\mathbf{x}}_{\ell})'$ & $\sum_{\ell=1}^{g}n_{\ell}-g$\tabularnewline
\hline 
Total: \quad 
$\mathbf{B}+\mathbf{W}=\sum_{\ell=1}^{g}\sum_{j=1}^{n_{\ell}}(\mathbf{x}_{\ell j}-\bar{\mathbf{x}})(\mathbf{x}_{\ell j}-\bar{\mathbf{x}})'$ & $\sum_{\ell=1}^{g}n_{\ell}-1$\tabularnewline
\hline 
\end{tabular}
\end{table}


## Wilks' Lamba $\Lambda^{*}$

* One test of 
\[
H_{0}:\tau_{1}=\tau_{2}=\cdots=\tau_{g}=0
\]
involves generalized variances. We reject $H_{0}$ if the ratio of generalized variances
\[
\Lambda^{*}=\frac{|\mathbf{W}|}{|\mathbf{B}+\mathbf{W}|}=\frac{\left|\sum_{\ell=1}^{g}\sum_{j=1}^{n_{\ell}}(\mathbf{x}_{\ell j}-\bar{\mathbf{x}}_{\ell})(\mathbf{x}_{\ell j}-\bar{\mathbf{x}}_{\ell})'\right|}{\left|\sum_{\ell=1}^{g}\sum_{j=1}^{n_{\ell}}(\mathbf{x}_{\ell j}-\bar{\mathbf{x}})(\mathbf{x}_{\ell j}-\bar{\mathbf{x}})'\right|}\,\,\,\text{is too small.}
\]  

* The quantity $\Lambda^{*}$ originally by Wilks corresponds to the equivalent form of the F-test of $H_{0}$: no treatment effects in the univariate case.  


## Wilks' Lamba $\Lambda^{*}$ (cont)

* Wilks $\Lambda^{*}$ can also be expressed as a function of the eigenvalues
$\hat{\lambda}_{1},\hat{\lambda}_{2},\ldots,\hat{\lambda}_{s}$ of
$\mathbf{W}^{-1}\mathbf{B}$ as
\[
\Lambda^{*}=\Pi_{i=1}^{s}\left(\frac{1}{1+\hat{\lambda}_{i}}\right),\,\,\text{where\,\ }s=\min(p,g-1)=rank(B)
\]


##  Distribution of Wilks' Lambda, $\Lambda^{*}$

\begin{table}
\centering
\begin{tabular}{|c|c|c|}
\hline 
$\begin{array}{c}
\text{No. of}\\
\text{variables}
\end{array}$ & $\begin{array}{c}
\text{No. of}\\
\text{groups}
\end{array}$ & $\begin{array}{c}
\text{Sampling distribution for}\\
\text{multivariate normal data}
\end{array}$\tabularnewline
\hline 
\hline 
$p=1$  & $g\geq2$ & $\left(\frac{\sum n_{\ell}-g}{g-1}\right)\left(\frac{1-\Lambda^{*}}{\Lambda^{*}}\right)\sim F_{g-1,\sum n_{\ell}-g}$\tabularnewline
\hline 
$p=2$ & $g\geq2$ & $\left(\frac{\sum n_{\ell}-g-1}{g-1}\right)\left(\frac{1-\sqrt{\Lambda^{*}}}{\sqrt{\Lambda^{*}}}\right)\sim F_{2(g-1),2(\sum n_{\ell}-g-1)}$\tabularnewline
\hline 
$p\geq1$ & $g=2$ & $\left(\frac{\sum n_{\ell}-p-1}{p}\right)\left(\frac{1-\Lambda^{*}}{\Lambda^{*}}\right)\sim F_{p,\sum n_{\ell}-g-1}$\tabularnewline
\hline 
$p\geq1$ & $g=3$ & $\left(\frac{\sum n_{\ell}-p-2}{p}\right)\left(\frac{1-\sqrt{\Lambda^{*}}}{\sqrt{\Lambda^{*}}}\right)\sim F_{2p,2(\sum n_{\ell}-p-2)}$\tabularnewline
\hline 
\end{tabular}
\end{table}


## Distribution of Wilks' Lambda, $\Lambda^{*}$ (cont)

For other cases and large sample sizes, $\sum_{\ell}n_{\ell}=n$ large,
we reject $H_{0}$ at significance level $\alpha$ if 
\begin{align*}
-\left(-n-1-\frac{(p+g)}{2}\right)\ln\Lambda^{*} & >\chi_{p(g-1)}^{2}(\alpha)\,\,\,\text{or}\\
\Lambda^{*}<exp & \left[-\left(-n-1-\frac{(p+g)}{2}\right)^{-1}\chi_{p(g-1)}^{2}(\alpha)\right]
\end{align*}


## Rootstock Data

The data contains four dependent variables as follows:

* six different rootstocks (Tree Number)
* trunk girth at four years (mm $\times$ 100)    
* extension growth at four years (m)  
* trunk girth at 15 years (mm $\times$ 100)  
* weight of tree above ground at 15 years (lb $\times$ 1000)

```{r}
library(ACSWR) # data is in package ACSWR
data("rootstock")
colnames(rootstock) <- c("Tree.Num", "Girth.4y",
  "Growth.4y", "Girth.15y", "WgtAbvGrnd.15y")
```


##

```{r}
head(rootstock)
```



## Setting up the data

The `manova()` function in R accepts formula interface `y ~ x`, where $y$ is the matrix of dependent variables (measurement value) and $x$ as the independent factor variable (population tree number).

```{r}
dep.variable <- as.matrix(rootstock[, 2:5])
ind.variable <- as.factor(rootstock[, 1])
rootstock.model <- manova( dep.variable ~ ind.variable)
```


## Wilks' MANOVA Test

```{r}
summary(rootstock.model, test = "W") # argument test = W
```

The resultant MANOVA model reports a Wilks' test statistic of 0.154 and a p-value below 0.05, thus $H_0$ is rejected and it is concluded there are significant differences in the means measurements of the six different rootstocks.


## Equivalent Test Statistics

In the context of random samples from several populations, the multivariate
tests are based on the matrices

\[
\mathbf{B}=\sum_{\ell=1}^{g}n_{\ell}(\bar{\mathbf{x}}_{\ell}-\bar{\mathbf{x}})(\bar{\mathbf{x}}_{\ell}-\bar{\mathbf{x}})'\,\,\,\text{and}\,\,\,\mathbf{W}=\sum_{\ell=1}^{g}\sum_{j=1}^{n_{\ell}}(\mathbf{x}_{\ell j}-\bar{\mathbf{x}}_{\ell})(\mathbf{x}_{\ell j}-\bar{\mathbf{x}}_{\ell})'
\]

We have used
\[
\text{Wilks lambda statistic}\,\,\,\Lambda^{*}=\frac{|\mathbf{W}|}{|\mathbf{B}+\mathbf{W}|}
\]
which is equivalent to the likelihood ratio test. 



## Other Multivariate Statistics

Three other multivariate test statistics are regularly included in the output of statistical packages
\begin{align*}
\text{Lawley-Hotelling Trace} & =tr[\mathbf{B}\mathbf{W}^{-1}]\\
\text{Pillai trace} & =tr[\mathbf{B}(\mathbf{B}+\mathbf{W})^{-1}]\\
\text{Roy's largest root} & =\text{maximum eigenvalue of}\,\,\,\,\mathbf{W}(\mathbf{B}+\mathbf{W})^{-1}
\end{align*}


## Equivalent Test Statistics

* All four of these tests appear to be nearly equivalent for extremely large samples.  

* For moderate sample sizes, all comparisons are based on what is necessarily a limited number of cases studied by simulation.  

* From the simulations reported, the first three tests have similar power, while the last, Roy's test, behaves differently.

* Its power is best only when there is a single nonzero eigenvalue and, at the same time, the power is large.


##

* There is also some suggestion that Pillai's trace is slightly more robust against nonnormality.  

* All four statistics apply in the two-way setting and in even more complicated MANOVA.  

* When, and only, when the multivariate tests signals a difference, or departure from the null hypothesis, do we probe deeper.


## Pillai's Statistic

```{r}
summary(rootstock.model)  # default output
```

## Hotelling-Lawley's Statistic

```{r}
summary(rootstock.model, test = "H")
```


## Roy's Statistics

```{r}
summary(rootstock.model, test = "R")
```


## Post Hoc Tests: 4 ANOVA's on the four tree measurements

```{r}
summary(aov(dep.variable ~ ind.variable))[1:2]
```

##

```{r}
summary(aov(dep.variable ~ ind.variable))[3:4]
```

Except for trunk girth at four years, there are significant differences in the means of rootstock measurements amongst the six groups.
